Assimilating disturbance: Toward real-time monitoring and forecasting

### 2023-03-31 methods.Rmd
unit change on obs error
Google Drive track changes

@@

The Ensemble Kalman Filter (EnKF) [@Evensen2009; @Evensen2009a] is a popular data assimilation algorithm that we use both to demonstrate the challenges conventional data assimilation approached have with capturing discrete disturbances and as our jumping off point for developing more flexible methods. Like most data assimilation algorithms the EnKF involves two cyclic steps: a Forecast step where a model is used to probabalistically predict the system state (and any other outputs) forward in time, and an Analysis step where new observations are used to update system state variables (e.g. carbon pools) and sometimes model parameters. These updated variables are then used as the initial conditions for the next Forecast round.

Within the EnKF, the probabalistic Forecast step is performed using an ensemble of model runs that vary in at least initial conditions, but potentially also model parameters, drivers, and process error. The ensemble forecast approach is easy to apply to any model, is particularly good at capturing nonlinear predictions, and unlike many other approaches does not require solving for the adjoint of the model. In EnKF the forecast, $f$, at any point in time, $t$, is then summarized in terms of a vector of means across ensemble members, $\mu_{f,t}$ and the sample covariance, $P_{f,t}$. In our example we are interested in the three state variables in the VSEM, ${C_l,C_w,C_s}$, at a single site and assume an ensemble size of 200. Therefore $\mu_{f,t}$ is a vector of length 3 and $P_{f,t}$ is a $3 \times 3$ covariance matrix. If we were making a spatial forecast at $n_s$ locations, $\mu_{f,t}$ would be a vector of length $3 n_s$ and $P_{f,t}$ would be a $3 n_s \times 3 n_s$ covariance matrix that would capture the full set of correlations across both variables and space. An important difference between data assimilation and spatial statistics is that in data assimilation the spatial covariance matrix comes from (1) the mechanistic understanding embedded in our process model and (2) the spatial covariance in the forecast inputs (initial conditions, parameters, drivers, process error), rather than an empirical semivariogram.

@@

While this fusion of forecast and new data is an incredibly powerful concept, as alluded to in the _Introduction_ there are times that the Kalman Normal-Normal "nudge" produces results that don't always make sense. Consider as a simple example a simulated data experiment where LAI growth follows a simple stochastic logistic model with growth rate $r$=0.1, equilibrium $K$=6, process error standard deviation 0.25, and observation error standard deviation 0.75. Under normal conditions, EnKF would have no trouble nudging a LAI forecast back on track every time a new observation is made, but what if a discrete disturbance event occurs knocking LAI from steady-state down to $1.0 \pm 0.5$? In this case the compromise between the forecast ($LAI=6$) and observations ($LAI=1$) produces a compromise ($LAI \approx 5$ that is a poor representation of the state of the system (Figure \@ref(fig:enkf)). Furthermore, as the true LAI slowly recovers upward, the EnKF keeps gradually nudging LAI downward until the two converge. Not only is the assimilated state wrong, but so is our estimate of the process -- a discrete disturbance and rapid recovery is behaving like gradual degradation, which will lead to incorrect inferences about related processes (litter, soil C, NPP, NEE, ET, albedo, etc.). Not also that including a disturbance probability in the process model does not solve this problem. Assuming the disturbance probability is low (e.g. a background rates of <1%), at every point in time the forecast would include a small number of disturbed ensemble members, but the overwhelming majority would be undisturbed. Therefore $\mu_{f,t}$ will be relatively unchanged and $P_{f,t}$ would see a small increase in variance. The fundamental problem is that even though the true forecast is now bimodal (a large peak at the undisturbed state and a much smaller peak at the disturbed state), the EnKF is still representing this as a unimodal Normal. Indeed, if the you increased the disturbance probability substantially (e.g. 50%), the EnKF would treat this as a unimodal forecast half way between the disturbed and undisturbed states, a place where no ensemble members are present (Figure \@ref(fig:concept), top).

@@

```{r concept, echo=FALSE, message = FALSE,error=FALSE, fig.asp=1, fig.cap="Top: In a conventional EnKF, a multimodal forecast (grey histogram) is approximated by a unimodal Normal (green), such that when a low probability disturbance is actually observed (red) the posterior (orange) does not capture this event. Bottom: In the multinomial filter the same multimodal forecast is represented by a mixture of normals (green and dark red) each weighted by their prior (forecast) probabulities. The posterior (blue), updates the mean and variance of each mode as well as their relative probabilities, in this case capturing the high probability that a disturbance occurred."}
library(rjags)
par(mfrow=c(2,1))
## params
set.seed(1066)
p.dist = 0.05
n0 = 1
nIC = 6
r = .1
K = 6
n.sd = 0.25
NE = 1000
obs=1
obs.sd = 0.75

@@

To address this fundamental flaw in the Kalman Analysis, we need to build into the Analysis step both the possibility for ensemble forecasts to be multimodel, and the ability for the data assimilation to jump discretely between modes all-at-once rather than through gradual nudges. To begin, let's use $X$ to represent the latent (unobserved) "true" state of the system. Furthermore, we will consider a process-model that is capable of predicting $k$ different categorical outcomes (land cover classes, types of disturbance, etc.) each of which has its own prediction for the latent continuous state variables of interest, which we'll refer to as $X^L_{k,t}$. Similar to the EnKF we'll assume that the ensemble prediction for any one of these categories can be represented as a multivariate Normal based on the ensemble sample mean and covariance:

@@

However, because we have $k$ categories we end up with a multimodel prediction. Furthermore, these modes need not be distinct; it is perfectly fine in concept for these MVNs to be overlapping. In practice they more they overlap the harder it is to identify which mode you are in, but as we will see this categorical assignment will be treated probabalistically. Indeed, to be able to represent our multimodel prediction as a single probability distribution we need to know the _forecasted_ probability of being in each category, which we'll take as known and represent using the vector $\vec{p_t}$, giving us a weighted MVN mixture as our forecast.

@@

$$X_t = X^L_t \vec{\rho_t}$$
which for any individual draw of $\vec{\rho_t}$ assigns $X_t$ that specific $X^L_{k,t}$. In practice, we will use Bayes' Theorem to integrate over the uncertainties in both our categorical assignment and continuous state variables, producing posterior Analysis probability distributions for both $X_t$ and $\vec{\rho_t}$. Unlike in the EnKF, where conjugacy allows us to calculate an analytical solution for the posterior, there is not simple analytical solution for the joint posterior distribution of $X_t$ and $\vec{\rho_t}$. In this case we will use well-developed Bayesian Markov Chain Monte Carlo (MCMC) numerical algorithms to sample from the posterior distribution. Specifically, numerical analyses in this paper were conducted using JAGS v4.2.0 [@Plummer03] accessed from R v4.0.3 [@Rcore] using the rjags package v4.10 [@rjags].

@@

@@

Our forth experiment was similar to the third, but extended the single-site, multi-disturbance, multi-pool scenario to one where one disturbance type included a switch in PFT. Specifically PFT 2 was identical to the VSEM default, but PFT 1 was parameterized with a stem residence time that was increased by 10x to create a more tree-like PFT. In addition, we established a transition matrix where disturbance 1 (stand clearing) transitions from PFT1 to PFT2, while disturbance 2 (defoliation) keeps the current PFT is. As with the previous scenario, the prescribed disturbance in the pseudodata was type 1 (stand clearing, transition from forest to shrub).

@@

@@

Our study region was a Landsat tile over central Oregon, USA located between 120 47'W and 123 7'W longitude and 43 45'N and 45 '4'N latitude. For this tile we extracted data from three derived Landtrendr datasets (http://emapr.ceoas.oregonstate.edu/pages/data/viz/index.html, tile h03v04):  biomass, cover, and disturbance. All three products were 30m resolution and annual timestep;biomass and cover extended from 1990-2017, while disturbance extended from 1984-2011 and a nearest neighbor algorithm was used to regrid the disturbance data to be the same as biomass and cover. The biomass product calibrated Landtrendr-processed Landsat data against field plot data using the gradient nearest neighbor (GNN) methodology to produce an annual biomass estimates [@Kennedy2018]. The cover product trained a spatiotemporal exploratory model (STEM), which uses an ensemble of regression trees defined within spatially overlapping support sets, against the 2001 U.S. National Land Cover Database (NLCD) to produce annual land-cover maps for Washington, Oregon, and California [@HOOPER2018]. To similify the analysis we focused on a single vegetation type, evergreen forest, which is the dominant vegetation cover in the region and consists primarily of Ponderosa pine (_Pinus ponderosa_). 
The disturbance product **need text from Robert** [@KENNEDY2010]. 
To simplify analyses we aggregated the disturbance product into five classes: undisturbed, cut, fire, pest, and other.

@@

Prior to calibration we set the leaf area ratio (LAR) to 15.7 m2/kgC and leaf residence time (tauL) to 4.7 years based on data from the BADM and left the light extinction coefficient (KEXT) at its default (0.5). The remaining fit parameters were the light use efficiency (LUE), the fraction of GPP allocated to autotrophic respiration (GAMMA), the fraction of NPP allocated to leaves (Av), the wood and soil C residence times (tauW, tauS), and the residual standard deviations for each of our data constraints (sd_NEE, sd_leaf, sd_wood). The fit was done taking a Maximum Likelihood approach using base R's `optim` function. Finally, because the process error within a dynamical data assimilation system is not identical to the calibration residual error, we estimated the leaf and wood process errors (Ql, Qw) based on the difference between the predicted and observed net annual changes in each of these pools. Because there was not an equivalent annually-observed soil carbon dataset, the soil carbon process error (Qs) was estimated based on the predicted mean annual rate of change under the simplifying assumption that the true soil carbon is in steady-state.

@@

For each location we next extracted PAR driver data from the Daymet product [CITE] using the daymetr R package [@daymetr]. For each location the carbon pools were initialized using a multi-step process to account for the lack of leaf and soil C data at a landscape scale. To start we used the first time point in each biomass time-series as the initial condition for the wood pool. Next, using the US-ME2 initial conditions we performed a >10k year spin-up, and then sampled carbon pool ensemble members from the second half of this time-series. Next, for each ensemble member we imposed at disturbance that reduced wood and leaf biomass to 1% of their equilibrum values, then simulated a 200-year recovery trajectory. Finally, for each ensemble member we selected the stand age that minimized the squared error between the modeled biomass and the initial biomass, and then selected the leaf and soil C pools at that stand age. This results is an ensemble of leaf, wood, and soil carbon pools with a covariance structure that is consistent with the calibrated model's dynamics.

For the assimilation experiment, each site was assimilated separately. In the initial round of experiments the observation error variance was set to $26.5\: kg^2/m^4$ based on [@Kennedy2018] Figure 2. Post-assimilation, we then extracted a number of metrics to assess algorithm performance: the posterior probability of disturbance in 2004, the posterior biomass pool size before disturbance, after disturbance, and at the end of the time-series (2017).
One thing we wanted to assess with the Multinomial filter is the extent of filter divergence, the situation where the model in a data assimilation algorithm becomes so confident in itself that assimilated observations have little impact on the posterior, which then diverges from observations.
As a metric to assess the prevalence of filter divergence, we calculated the difference between the last observed data point and the posterior mean of the corresponding Analysis, and then divided this by half the width of the Analysis 95% confidence interval.
If the confidence interval is symmetric, then any Analysis mean that falls in the confidence interval will score between -1 and 1, while values outside this range are suggestive of filter divergence.

@@

Finally, based on preliminary results we performed a series of additional assimilation experiments where we manually altered the observation error, R, or process error, Q, to assess the sensitivity of our results. Table (\@ref(tab:QRexp)) lists the multipliers that were used to adjust R and Q. The skill of these assimilation experiments were assessed the same as the original. In addition, changes in skill and filter divergence across experiments were assessed as a function of R and Q focusing on two response variables: (1) the filter coverage, defined as the proportion of sites where our divergence metric falls between -1 and 1 (i.e., where the difference between the observation and Analysis is less than half the width of the Analysis 95% CI); and (2) the detection threshold, defined as the change in AGB where the probability of detecting a disturbance is 50% as inferred by the GLM of disturbance size versus probability of disturbance.

>>>>

**TODO Dongchen: consider adding a conceptual diagram to explain the framework and data**

@@

The Ensemble Kalman Filter (EnKF) [@Evensen2009; @Evensen2009a] is a popular data assimilation algorithm that we use both to demonstrate the challenges conventional data assimilation approaches have with capturing discrete disturbances and as our jumping off point for developing more flexible methods. Like most data assimilation algorithms the EnKF involves two cyclic steps: a Forecast step where a model is used to probabilistically predict the system state (and any other outputs) forward in time, and an Analysis step where new observations are used to update system state variables (e.g. carbon pools) and sometimes model parameters. These updated variables are then used as the initial conditions for the next Forecast round.

Within the EnKF, the probabilistic Forecast step is performed using an ensemble of model runs that vary in at least initial conditions, but potentially also model parameters, drivers, and process error. The ensemble forecast approach is easy to apply to any model, is particularly good at capturing nonlinear predictions, and unlike many other approaches does not require solving for the adjoint of the model. In EnKF the forecast, $f$, at any point in time, $t$, is then summarized in terms of a vector of means across ensemble members, $\mu_{f,t}$ and the sample covariance, $P_{f,t}$. In our example we are interested in the three state variables in the VSEM, ${C_l,C_w,C_s}$, at a single site and assume an ensemble size of 200. Therefore $\mu_{f,t}$ is a vector of length 3 and $P_{f,t}$ is a $3 \times 3$ covariance matrix. If we were making a spatial forecast at $n_s$ locations, $\mu_{f,t}$ would be a vector of length $3 n_s$ and $P_{f,t}$ would be a $3 n_s \times 3 n_s$ covariance matrix that would capture the full set of correlations across both variables and space. An important difference between data assimilation and spatial statistics is that in data assimilation the spatial covariance matrix comes from (1) the mechanistic understanding embedded in our process model and (2) the spatial covariance in the forecast inputs (initial conditions, parameters, drivers, process error), rather than an empirical semivariogram.

@@

While this fusion of forecast and new data is an incredibly powerful concept, as alluded to in the _Introduction_ there are times that the Kalman Normal-Normal "nudge" produces results that do not make sense. Consider as a simple example a simulated data experiment where LAI growth follows a simple stochastic logistic model with growth rate $r$=0.1, equilibrium $K$=6, process error standard deviation 0.25, and observation error standard deviation 0.75. Under normal conditions, EnKF would have no trouble nudging a LAI forecast back on track every time a new observation is made, but what if a discrete disturbance event occurs knocking LAI from steady-state down to $1.0 \pm 0.5$? In this case the compromise between the forecast ($LAI=6$) and observations ($LAI=1$) produces a compromise ($LAI \approx 5$ that is a poor representation of the state of the system (Figure \@ref(fig:enkf)). Furthermore, as the true LAI slowly recovers upward, the EnKF keeps gradually nudging LAI downward until the two converge. Not only is the assimilated state wrong, but so is our estimate of the process -- a discrete disturbance and rapid recovery is behaving like gradual degradation, which will lead to incorrect inferences about related processes (litter, soil C, NPP, NEE, ET, albedo, etc.). It is also worth noting that including a disturbance probability in the process model does not, by itself, solve this problem. Assuming the disturbance probability is low (e.g. a background rate of <1%), at every point in time the forecast would include a small number of disturbed ensemble members, but the overwhelming majority would be undisturbed. Therefore $\mu_{f,t}$ will be relatively unchanged and $P_{f,t}$ would see a small increase in variance. The fundamental problem is that even though the true forecast is now bimodal (a large peak at the undisturbed state and a much smaller peak at the disturbed state), the EnKF is still representing this as a unimodal Normal centered around the undisturbed state. Indeed, if the disturbance probability were increased substantially (e.g. 50%), the EnKF would treat this as a unimodal forecast halfway between the disturbed and undisturbed states, a place where no ensemble members are present (Figure \@ref(fig:concept), top).

@@

** TODO from zoey for above figure: Labels could be more informative (i.e. EnKF = conventional and Multinomial = improved approach). Icon for "data" should match the icons rather than a line. Is the P(dist) bound at 0,1? I barely noticed the black line at first, it could maybe go in a second panel below with a shared x-axis and a distinct y-axis.**

** TODO from zoey for below figure: I think this figure could be more streamlined to better emphasize the difference between the filters. Why change the assimilation color from orange to blue in this figure?   The histogram, to me, seems more intuitive for displaying the data values rather than the forecasted values. The second panel doesn't have "forecast" in the color labels which also adds to confusion. The X axis could communicate the low-LAI = disturbed, high-LAI = undisturbed, without relying on the colors/legend. X axis for this and the Y axis for the above figure could just be "disturbance magnitude" **

```{r concept, echo=FALSE, message = FALSE,error=FALSE, fig.asp=1, fig.cap="Top: In a conventional EnKF, a multimodal forecast (gray histogram) is approximated by a unimodal Normal (green), such that when a low probability disturbance is actually observed (red) the posterior (orange) does not capture this event. Bottom: In the multinomial filter the same multimodal forecast is represented by a mixture of normals (green and dark red) each weighted by their prior (forecast) probabilities. The posterior (blue), updates the mean and variance of each mode as well as their relative probabilities, in this case capturing the high probability that a disturbance occurred."}
library(rjags)
par(mfrow=c(2,1))
## params
set.seed(1066)
p.dist = 0.05
n0 = 1
nIC = 6
r = .1
K = 6
n.sd = 0.25
NE = 1000
obs=1
obs.sd = 0.75

@@

To address this fundamental flaw in the Kalman Analysis, we need to build into the Analysis step both the possibility for ensemble forecasts to be multimodal, and the ability for the data assimilation to jump discretely between modes all-at-once rather than through gradual nudges. To begin, let's use $X$ to represent the latent (unobserved) "true" state of the system. Furthermore, we will consider a process-model that is capable of predicting $k$ different categorical outcomes (land cover classes, types of disturbance, etc.) each of which has its own prediction for the latent ($L$) continuous state variables of interest, which we'll refer to as $X^L_{k,t}$. Similar to the EnKF we'll assume that the ensemble forecast for any one of these categories can be represented as a multivariate Normal,

@@

where $\vec{\mu}_{f,k,t}$ and $P_{f,k,t}$ are the ensemble forecast ($f$) sample mean and covariance for outcome $k$ at time $t$. However, because we have $k$ categories we end up with a multimodal prediction (i.e., a mixture of MVNs). Furthermore, these modes need not be distinct; it is perfectly fine in concept for these MVNs to be overlapping. In practice the more they overlap the harder it is to identify which mode you are in, but as we will see this categorical assignment will be treated probabilistically. Indeed, to be able to represent our multimodal prediction as a single probability distribution we need to know the _forecasted_ probability of being in each category, which we'll take as known and represent using the vector $\vec{p_t}$, giving us a weighted MVN mixture as our forecast.

@@

$$X_t = X^L_t \vec{\rho_t}$$
which for any individual draw of $\vec{\rho_t}$ assigns $X_t$ that specific $X^L_{k,t}$. In practice, we will use Bayes' Theorem to integrate over the uncertainties in both our categorical assignment and continuous state variables, producing posterior Analysis probability distributions for both $X_t$ and $\vec{\rho_t}$. Unlike in the EnKF, where conjugacy allows us to calculate an analytical solution for the posterior, there is no simple analytical solution for the joint posterior distribution of $X_t$ and $\vec{\rho_t}$. In this case we will use well-developed Bayesian Markov Chain Monte Carlo (MCMC) numerical algorithms to sample from the posterior distribution. Specifically, numerical analyses in this paper were conducted using JAGS v4.2.0 [@Plummer03] accessed from R v4.0.3 [@Rcore] using the rjags package v4.10 [@rjags].

** TODO: Adjust fig 2 to make colors consistent between panels and with other figures**

@@

** TODO: Check equation formatting **

@@

Our fourth experiment was similar to the third, but extended the single-site, multi-disturbance, multi-pool scenario to one where one disturbance type included a switch in PFT. Specifically PFT 2 was identical to the VSEM default, but PFT 1 was parameterized with a stem residence time that was increased by 10x to create a more tree-like PFT. In addition, we established a transition matrix where disturbance 1 (stand clearing) transitions from PFT1 to PFT2, while disturbance 2 (defoliation) keeps the current PFT is. As with the previous scenario, the prescribed disturbance in the pseudodata was type 1 (stand clearing, transition from forest to shrub).

@@

** TODO: TRY AGAIN AT GETTING MAPS ON SAME PROJECTION **

@@

Our study region was a portion of central Oregon, USA that includes the Ameriflux tower used later to initialize the VSEM (see next section).  We used biomass, cover, and disturbance data derived from time-series Landsat imagery processed with the LandTrendr algorithm under a separate project.  Available on a tile-basis (http://emapr.ceoas.oregonstate.edu/pages/data/viz/index.html), we downloaded these data for tile h03v04, which corresponds to the region between 120 47'W and 123 7'W longitude and 43 45'N and 45 '4'N latitude. All three products were 30m resolution and annual timestep;biomass and cover extended from 1990-2017, while disturbance extended from 1984-2011 and a nearest neighbor algorithm was used to regrid the disturbance data to be the same as biomass and cover. The biomass product was constructed by linking Landtrendr-processed Landsat data with field plot data using the gradient nearest neighbor (GNN) methodology [@Ohmann2002] to produce annual biomass estimates [@Kennedy2018]. The cover product trained a spatiotemporal exploratory model (STEM), which uses an ensemble of regression trees defined within spatially overlapping support sets, against the 2001 U.S. National Land Cover Database (NLCD) to produce annual land-cover maps for Washington, Oregon, and California [@HOOPER2018]. To simplify the analysis we focused on a single vegetation type, evergreen forest, which is the dominant vegetation cover in the region and consists primarily of Ponderosa pine (_Pinus ponderosa_). 
The disturbance product linked LandTrendr-derived disturbance information [@KENNEDY2010] with photo-interpreted labels of disturbance to attribute drivers to each disturbance event [@KENNEDY2015].  
To simplify analyses we aggregated the disturbance product into five classes: undisturbed, cut, fire, pest, and other.

@@

Prior to calibration we set the leaf area ratio (LAR) to 15.7 m2/kgC and leaf residence time (tauL) to 4.7 years based on data from the BADM and left the light extinction coefficient (KEXT) at its default (0.5). The remaining fit parameters were the light use efficiency (LUE), the fraction of GPP allocated to autotrophic respiration (GAMMA), the fraction of NPP allocated to leaves (Av), the wood and soil C residence times (tauW, tauS), and the residual standard deviations for each of our data constraints (sd_NEE, sd_leaf, sd_wood). The fit was done taking a Maximum Likelihood approach using base R's `optim` function. Finally, because the process error within a dynamical data assimilation system is not identical to the calibration residual error, we estimated the leaf and wood process errors (Ql, Qw) based on the difference between the predicted and observed net annual changes in each of these pools. Because there was not an equivalent annually-observed soil carbon dataset, the soil carbon process error (Qs) was estimated based on the predicted mean annual rate of change under the simplifying assumption that the true soil carbon is in steady-state (i.e. any change in modeled soil C was assumed to be an error).

@@

For each location we next extracted PAR driver data from the Daymet product [CITE] using the daymetr R package [@daymetr]. For each location the carbon pools were initialized using a multi-step process to account for the lack of leaf and soil C data at a landscape scale. To start we used the first time point in each biomass time-series as the initial condition for the wood pool. Next, using the US-ME2 initial conditions we performed a >10k year spin-up, and then sampled carbon pool ensemble members from the second half of this time-series. Next, for each ensemble member we imposed at disturbance that reduced wood and leaf biomass to 1% of their equilibrium values, then simulated a 200-year recovery trajectory. Finally, for each ensemble member we selected the stand age that minimized the squared error between the modeled biomass and the initial biomass, and then selected the leaf and soil C pools at that stand age. This results is an ensemble of leaf, wood, and soil carbon pools with a covariance structure that is consistent with the calibrated model's dynamics.

For the assimilation experiment, each site was assimilated separately. In the initial round of experiments the observation error was set to a standard deviation of $4.2\: kg/m^2$ based on [@Kennedy2018] Figure 2. Post-assimilation, we then extracted a number of metrics to assess algorithm performance: the posterior probability of disturbance in 2004, the posterior biomass pool size before disturbance, after disturbance, and at the end of the time-series (2017).
One thing we wanted to assess with the Multinomial filter is the extent of filter divergence, the situation where the model in a data assimilation algorithm becomes so confident in itself that assimilated observations have little impact on the posterior, which then diverges from observations.
As a metric to assess the prevalence of filter divergence, we calculated the difference between the last observed data point and the posterior mean of the corresponding Analysis, and then divided this by half the width of the Analysis 95% confidence interval.
If the confidence interval is symmetric, then any Analysis mean that falls in the confidence interval will score between -1 and 1, while values outside this range are suggestive of filter divergence.

@@

Finally, based on preliminary results we performed a series of additional assimilation experiments where we manually altered the observation error, R, or process error, Q, to assess the sensitivity of our results. Table (\@ref(tab:QRexp)) lists the multipliers that were used to adjust R and Q. The skill of these assimilation experiments were assessed the same as the original. In addition, changes in skill and filter divergence across experiments were assessed as a function of R and Q focusing on two response variables: (1) the filter coverage, defined as the proportion of sites where our divergence metric falls between -1 and 1 (i.e., where the difference between the observation and Analysis is less than half the width of the Analysis 95% CI); and (2) the detection threshold, defined as the change in AGB where the probability of detecting a disturbance is 50% as inferred by the GLM of disturbance size versus probability of disturbance detection.
